### 2023-12-30 talk/aprm24.tex
aprm talk submitted

\item We introduce the {\bf iterated block particle filter}, currently the most powerful algorithm available in the \texttt{spatPomp} R package.

@@

    \begin{itemize}
    \item  Iterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do plug-and-play likelihood-based inference:

@@

\begin{frame}{COVID in 375 cities in China, 10-23 January, 2020}

  \begin{itemize}
  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness
    (Li et al, {\it Science}, May 2020).
  \item SEIR (susceptible-exposed-infected-removed) model with asymptomatics, reporting delay, and coupling based on cell phone data.
  \item Li et al (2020) used iterated EnKF for inference.
  \item The time interval covers the initial China lockdown.
  \item Here, we summarize our re-analysis of this model and data, recently posted on arXiv.
%% see README for R code
%   \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
%   \item{} 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
%   \item{} 0 cases reported for all cities during Jan 10--15.

@@

      \item This motivated us to develop an IBPF for parameter estimation.

@@

@@

%% \begin{frame}{Technical issues with EnKF}

%%   \begin{itemize}
%%   \item EnKF is based on a continuous Gaussian approximation.
%%   \item Log-likelihoods with respect to counting measure are never positive, and cannot properly be compared with log-likelihoods corresponding to continuous densities.
%%     \item For data with many zeros, the unbounded EnKF likelihood can substantially bias the MLE.
%%     \item Adapting EnKF for count data is non-trivial (Katzfuss et al, {\it JASA}, 2019).
%% %    \item Li et al avoided the issue by using a ``heuristic observation error variance'' within EnKF. The corresponding ``log-likelihood'' is not the log-likelihood of any model. In particular, it cannot validly be compared againt simple statistical benchmarks.
%% %  \item Data and models will never be perfect. We want to reveal, rather than conceal, limitations.
%%   \end{itemize}

%%   \end{frame}

%% \begin{frame}{Future work}

%%   \newcommand\futuresep{\vspace{5mm}}

%%   \begin{myitemize}

%%   \item We anticipate continued progress in modeling and inference for nonlinear spatiotemporal partially observed stochastic dynamic systems.

%%     \futuresep

%%   \item As long as the \texttt{spatPomp} package facilitates data analysis, we expect continuing package development.

%% %    \futuresep

%%  %   \item Questions and comments?

%% \end{myitemize}

%% \end{frame}

@@

\nocite{asfaw23arxiv,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides21,ionides22,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler23}

>>>>

\item We introduce the {\bf iterated block particle filter}, currently the most effective algorithm in the \texttt{spatPomp} R package.

@@

    \begin{itemize}
    \item  Iterated particle filtering via \texttt{mif2} in the R package \texttt{pomp} enables Masters-level statisticians to do plug-and-play likelihood-based inference for nonlinear, non-Gaussian, partially observed dynamic systems:

@@

\begin{frame}{COVID in 373 cities in China, Jan 10 to Feb 8, 2020}

  \begin{itemize}
  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness
    (Li et al, {\it Science}, May 2020).
  \item SEIR (susceptible-exposed-infected-removed) model with asymptomatics, reporting delay, and coupling based on cell phone data.
  \item Li et al (2020) used iterated EnKF for inference.
  \item The time interval covers the initial China lockdown.
  \item We present re-analysis of this model and data, recently posted on arXiv.
%% see README for R code
%   \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
%   \item{} 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
%   \item{} 0 cases reported for all cities during Jan 10--15.

@@

      \item This motivated us to develop an iterated BPF (IBPF) for parameter estimation.

@@

    \item Two recent case studies (Wheeler et al, 2023; Li et al, 2023) demonstrate data analysis using IBPF. Code and data are provided via R packages extending \code{spatPomp}.

@@

@@

\nocite{asfaw23arxiv,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides21,ionides22,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler23,li23}
